noam_beta_diversity

### 2024-09-06 noam_beta_diversity.Rmd
finished quad level beta dispersion models and figs

```{r quad level beta dispersion}
full_species_matrix <- as.matrix(data[ , -which(names(data) %in% meta_cols)])
full_species_matrix <- full_species_matrix/data$hard_cover # convert from percent to proportion
rownames(full_species_matrix) <- paste(data$year,"-", data$location, sep = "")
full_rows_to_remove <- rowSums(full_species_matrix) == 0 # remove rows with no algal species observed - thought this was done already
full_species_matrix <- full_species_matrix[!full_rows_to_remove,] 
full_species_matrix <- na.omit(full_species_matrix)
full_species_matrix <- asin(sqrt(full_species_matrix)) # arcsin squareroot transform
data$stratum <- paste(data$year,"-", data$location, sep = "") # for merging with bray matrix later
full_subset <- data[data$stratum %in% rownames(full_species_matrix),] # make sure rows are also removed from dataframe

brayfull <- vegdist(full_species_matrix, "bray") # create bray dissimilarity matrix
# full_quad_betadisper <- vegan::betadisper(brayfull, full_subset$hab_year)
full_subset$betadispersion <- full_quad_betadisper$distances[match(full_subset$stratum, names(full_quad_betadisper$distances))]

full_subset$betadispersion_trans <- sv_trans(full_subset$betadispersion, full_subset)

@@

macro_species_matrix <- as.matrix(macroalgae[ , -which(names(macroalgae) %in% meta_cols)])
macro_species_matrix <- macro_species_matrix/macroalgae$hard_cover # convert from percent to proportion
rownames(macro_species_matrix) <- paste(macroalgae$year,"-", macroalgae$location, sep = "")
macro_rows_to_remove <- rowSums(macro_species_matrix) == 0 # remove rows with no algal species observed - thought this was done already
macro_species_matrix <- macro_species_matrix[!macro_rows_to_remove,] 
macro_species_matrix <- na.omit(macro_species_matrix)
macro_species_matrix <- asin(sqrt(macro_species_matrix)) # arcsin squareroot transform
macroalgae$stratum <- paste(macroalgae$year,"-", macroalgae$location, sep = "") # for merging with bray matrix later
macro_subset <- macroalgae[macroalgae$stratum %in% rownames(macro_species_matrix),] # make sure rows are also removed from macroalgaeframe

braymacro <- vegdist(macro_species_matrix, "bray") # create bray dissimilarity matrix
# macro_quad_betadisper <- vegan::betadisper(braymacro, macro_subset$hab_year)
macro_subset$betadispersion <- macro_quad_betadisper$distances[match(macro_subset$stratum, names(macro_quad_betadisper$distances))]

>>>>

```{r quad level beta dispersion, fig.height = 8, fig.width = 12}
full_species_matrix <- as.matrix(data[ , -which(names(data) %in% meta_cols)])
full_species_matrix <- full_species_matrix/data$hard_cover # convert from percent to proportion
rownames(full_species_matrix) <- paste(data$year,"-", data$location, sep = "")
full_rows_to_remove <- rowSums(full_species_matrix) == 0 # remove rows with no algal species observed - thought this was done already
full_species_matrix <- full_species_matrix[!full_rows_to_remove,] 
full_species_matrix <- na.omit(full_species_matrix)
full_species_matrix <- asin(sqrt(full_species_matrix)) # arcsin squareroot transform
data$stratum <- paste(data$year,"-", data$location, sep = "") # for merging with bray matrix later
# full_subset <- data[data$stratum %in% rownames(full_species_matrix),] # make sure rows are also removed from dataframe
# 
# brayfull <- vegdist(full_species_matrix, "bray") # create bray dissimilarity matrix
# # full_quad_betadisper <- vegan::betadisper(brayfull, full_subset$hab_year)
# full_subset$betadispersion <- full_quad_betadisper$distances[match(full_subset$stratum, names(full_quad_betadisper$distances))]
# full_subset$betadispersion_trans <- sv_trans(full_subset$betadispersion, full_subset)
# 
# write_csv(full_subset, here("data", "quad_level_all_species_with_betadisper.csv"))

full_subset <- read.csv(here("data", "quad_level_all_species_with_betadisper.csv")) # read back in data from csv so you don't have to re-run

@@

emtrends(betadisper_quad_mod, pairwise ~ habitat , "year") # all different
betadisper_quad_emm <- emmip(betadisper_quad_mod, habitat ~ year , at = list(year = seq(2007,2023,1)), plotit = F, CIs = T)

ggplot()+
   geom_point(data = full_subset, aes(x = year, y = betadispersion_trans, colour = habitat), alpha = 0.1 ) +
   geom_line(data = betadisper_quad_emm, aes(x = year, y = inv_logit(yvar), colour = habitat), linewidth = 2) +
   geom_ribbon(data = betadisper_quad_emm, aes(x = year, ymin = inv_logit(LCL), ymax = inv_logit(UCL), fill = habitat), alpha = 0.3) +
   scale_colour_manual(values = habitat_colours) +
   scale_fill_manual(values = habitat_colours) +
   geom_segment(aes(y = 1, yend = 1, x = 2006, xend = 2010), linewidth = 2, alpha = 1, colour = "grey") +
   geom_vline(xintercept = 2010, linetype = 2, colour = "black", size = 1) +
   geom_vline(xintercept = 2019, linetype = 2, colour = "red", size = 1) +
   labs(y = "Beta dispersion\n", x = "Year") +
   theme_bw() + 
   theme(panel.grid.major = element_blank(), 
         panel.grid.minor = element_blank(),
         panel.background = element_blank(),
         plot.title = element_text(color = "grey20", size = 32, vjust = 0, face = "plain"),
         axis.line.y = element_line(colour = "black"),
         axis.text.y = element_text(color = "grey20", size = 28, hjust = .5, vjust = .5, face = "plain"),
         axis.title.y = element_text(color = "grey20", size = 32, hjust = .5, vjust = 0, face = "plain"),
         axis.line.x = element_line(colour = "black"),
         axis.text.x = element_text(color = "grey20", size = 28, angle = 90, vjust = 0.5, hjust=1, face = "plain"),
         axis.title.x = element_text(color = "grey20", size = 32, hjust = .5, vjust = 0, face = "plain"),
         strip.text.x = element_text(size = 32),
         legend.key.size = unit(2, "cm"), # increase legend size
         legend.text = element_text(size = rel(2)), # increase legend text
         legend.title = element_text(size = rel(2.25)),
         legend.position = 'bottom', # move legend to bottom
         legend.box = 'horizontal',
         legend.box.background = element_rect(colour = "black") )# make legend horizontal

macro_species_matrix <- as.matrix(macroalgae[ , -which(names(macroalgae) %in% meta_cols)])
macro_species_matrix <- macro_species_matrix/macroalgae$hard_cover # convert from percent to proportion
rownames(macro_species_matrix) <- paste(macroalgae$year,"-", macroalgae$location, sep = "")
macro_rows_to_remove <- rowSums(macro_species_matrix) == 0 # remove rows with no algal species observed - thought this was done already
macro_species_matrix <- macro_species_matrix[!macro_rows_to_remove,] 
macro_species_matrix <- na.omit(macro_species_matrix)
macro_species_matrix <- asin(sqrt(macro_species_matrix)) # arcsin squareroot transform
macroalgae$stratum <- paste(macroalgae$year,"-", macroalgae$location, sep = "") # for merging with bray matrix later
# macro_subset <- macroalgae[macroalgae$stratum %in% rownames(macro_species_matrix),] # make sure rows are also removed from macroalgaeframe
# 
# braymacro <- vegdist(macro_species_matrix, "bray") # create bray dissimilarity matrix
# # macro_quad_betadisper <- vegan::betadisper(braymacro, macro_subset$hab_year)
# macro_subset$betadispersion <- macro_quad_betadisper$distances[match(macro_subset$stratum, names(macro_quad_betadisper$distances))]
# macro_subset$betadispersion_trans <- sv_trans(macro_subset$betadispersion, macro_subset)
# write_csv(macro_subset, here("data", "quad_level_just_macroalgae_with_betadisper.csv"))
macro_subset <- read.csv(here("data", "quad_level_just_macroalgae_with_betadisper.csv"))

betadisper_macro_quad_mod <- glmmTMB(betadispersion_trans ~ year*habitat + (1|site), family = beta_family(link = "logit"), data = macro_subset)
summary(betadisper_macro_quad_mod) # fair bit of within site variance - inv_logit(0.02) = 0.5
qqnorm(residuals(betadisper_macro_quad_mod)) # a little lumpy but not bad
plot(residuals(betadisper_macro_quad_mod) ~ fitted(betadisper_macro_quad_mod)) # negative trend
Anova(betadisper_macro_quad_mod)

#                 Chisq Df Pr(>Chisq)    
# year           97.391  1  < 2.2e-16 ***
# habitat      1973.896  3  < 2.2e-16 ***
# year:habitat  511.722  3  < 2.2e-16 ***

emtrends(betadisper_macro_quad_mod, pairwise ~ habitat , "year") # backreef, outer10, and fringing don't differ
betadisper_quad_macro_emm <- emmip(betadisper_macro_quad_mod, habitat ~ year , at = list(year = seq(2007,2023,1)), plotit = F, CIs = T)

ggplot()+
   geom_point(data = macro_subset, aes(x = year, y = betadispersion_trans, colour = habitat), alpha = 0.1 ) +
   geom_line(data = betadisper_quad_macro_emm, aes(x = year, y = inv_logit(yvar), colour = habitat), linewidth = 2) +
   geom_ribbon(data = betadisper_quad_macro_emm, aes(x = year, ymin = inv_logit(LCL), ymax = inv_logit(UCL), fill = habitat), alpha = 0.3) +
   scale_colour_manual(values = habitat_colours) +
   scale_fill_manual(values = habitat_colours) +
   geom_segment(aes(y = 1, yend = 1, x = 2006, xend = 2010), linewidth = 2, alpha = 1, colour = "grey") +
   geom_vline(xintercept = 2010, linetype = 2, colour = "black", size = 1) +
   geom_vline(xintercept = 2019, linetype = 2, colour = "red", size = 1) +
   labs(y = "Beta dispersion\n", x = "Year") +
   theme_bw() + 
   theme(panel.grid.major = element_blank(), 
         panel.grid.minor = element_blank(),
         panel.background = element_blank(),
         plot.title = element_text(color = "grey20", size = 32, vjust = 0, face = "plain"),
         axis.line.y = element_line(colour = "black"),
         axis.text.y = element_text(color = "grey20", size = 28, hjust = .5, vjust = .5, face = "plain"),
         axis.title.y = element_text(color = "grey20", size = 32, hjust = .5, vjust = 0, face = "plain"),
         axis.line.x = element_line(colour = "black"),
         axis.text.x = element_text(color = "grey20", size = 28, angle = 90, vjust = 0.5, hjust=1, face = "plain"),
         axis.title.x = element_text(color = "grey20", size = 32, hjust = .5, vjust = 0, face = "plain"),
         strip.text.x = element_text(size = 32),
         legend.key.size = unit(2, "cm"), # increase legend size
         legend.text = element_text(size = rel(2)), # increase legend text
         legend.title = element_text(size = rel(2.25)),
         legend.position = 'bottom', # move legend to bottom
         legend.box = 'horizontal',
         legend.box.background = element_rect(colour = "black") )# make legend horizontal
```
